# Measures of diagnostic test accuracy {#sec-diagnostic_accuracy}


When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Evaluate a diagnostic test with dichotomous test result
-   Interpret the results
:::



## Research question

To estimate the diagnostic accuracy of digital mammography (index test) in the detection of breast cancer, using histopathology as a "gold standard" in women aged over 40 years, who are undergoing mammography for the evaluation of different symptoms related to breast diseases.



## Packages we need

We need to load the following packages:

```{r}
#| message: false

library(epiR)

library(here)
library(tidyverse)
```





## Contingency 2x2 table

Generally, an individual's disease status is a dichotomous variable; the individual either has the disease ($D_+$) or hasn't the disease ($D_-$) as defined by the reference standard (or "gold" standard). 
The diagnostic test can be measured with a dichotomous variable (e.g. presence or absence of breast abnormalities using an X-ray) or a continuous variable (e.g. fasting glucose level for diabetes diagnosis), that can be transformed to a dichotomous variable by choosing an optimal cut-off value (threshold) which distinguishes positive ($T_+$) from negative ($T_-$) test results.

If the diagnostic test gives a dichotomous result for each participant in a study the data can be tabulated in a 2 x 2 table of test result ($T_+$, $T_-$) versus true disease status ($D_+$, $D_-$). For example, the result of digital mammography test to diagnose breast cancer compared to the "gold" standard biopsy/surgery and histopathology in 1220 women with suspected breast cancer are following: 


```{r}
tb1 <- as.table(
  rbind(c(890, 110), c(20, 200))
  )

dimnames(tb1) <- list(
  Test = c("T+", "T_"),
  Disease = c("D+", "D-")
)

tb1
```

```{r}
epi.tests(tb1, digits = 3)
```




