# Paired t-test {#sec-paired_t_test}

A paired t-test is used to estimate whether the means of two related measurements are significantly different from one another.

Examples of paired study designs are:

1.  measurements collected **before and after** an intervention in an experimental study
2.  twins, husbands and wives, brothers and sisters, pairs of eyes
3.  **matched** cases and controls
4.  a **cross-over trial** in which each patient has two measurements on the variable, one while taking active treatment and one while taking placebo

When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Applying hypothesis testing
-   Compare two dependent (related) samples applying paired t-test
-   Interpret the results
:::

## Research question

The dataset *weight* contains the birth and discharge weight of 25 newborns. We might ask if the **mean difference** of the weight in birth and in discharge equals to zero or not. If the **differences** between the pairs of measurements are normally distributed, a paired t-test is the most appropriate statistical test.

::: {.callout-note icon="false"}
## Null hypothesis and alternative hypothesis

-   $H_0$: the mean difference of weight equals to zero ($\mu_{d} = 0$)
-   $H_1$: the mean difference of weight does not equal to zero ($\mu_{d} \neq 0$)
:::

## Packages we need

We need to load the following packages:

```{r}
#| message: false
#| warning: false

library(rstatix)
library(gtsummary)
library(here)
library(tidyverse)
library(report)
```

## Preraring the data

We import the data *weight* in R:

```{r}
#| warning: false

library(readxl)
weight <- read_excel(here("data", "weight.xlsx"))
```

We calculate the differences using the `mutate()` function:

```{r}
weight <- weight |> 
  mutate(dif_weight = discharge_weight - birth_weight)
```

```{r}
#| echo: false
#| label: fig-weight
#| fig-cap: Table with data from "weight" file.

DT::datatable(
  weight, extensions = 'Buttons', options = list(
    dom = 'tip',
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

We inspect the data:

```{r}
glimpse(weight) 
```

## Assumptions

::: {.callout-note icon="false"}
## Check if the following assumption is satisfied

The differences are **normally** distributed.
:::

## Explore the characteristics of distribution of differences

The distribution of the differences can be explored with appropriate plots and summary statistics.

**Graph**

We can explore the distribution of differences visually for symmetry with a density plot (a smoothed version of the histogram):

```{r}
#| warning: false
#| label: fig-hist1
#| fig-cap: Density plot of the weight differences.

weight |> 
  ggplot(aes(x = dif_weight)) +
  geom_density(fill = "#76B7B2", color="black", alpha = 0.2) +
  geom_vline(aes(xintercept=mean(dif_weight)),
             color="blue", linetype="dashed", size=1.4) +
  geom_vline(aes(xintercept=median(dif_weight)),
             color="red", linetype="dashed", size=1.2) +
  labs(x = "Weight difference") +
  theme_minimal() +
  theme(plot.title.position = "plot")
```

The above figure shows that the data are following an approximately symmetrical distribution. Note that the arithmetic mean (blue vertical dashed line) is very close to the median (red vertical dashed line) of the data.

**Summary statistics**

Summary statistics can also be calculated for the variables.

::: {#exercise-joins .callout-tip}
## Summary statistics

::: panel-tabset
## dplyr

We can utilize the `across` function to obtain the results across the three variables simultaneously:

```{r}
#| warning: false

summary_weight <- weight |> 
  dplyr::summarise(across(
    .cols = c(dif_weight, birth_weight, discharge_weight), 
    .fns = list(
      n = ~n(),
      na = ~sum(is.na(.)),
      min = ~min(., na.rm = TRUE),
      q1 = ~quantile(., 0.25, na.rm = TRUE),
      median = ~quantile(., 0.5, na.rm = TRUE),
      q3 = ~quantile(., 0.75, na.rm = TRUE),
      max = ~max(., na.rm = TRUE),
      mean = ~mean(., na.rm = TRUE),
      sd = ~sd(., na.rm = TRUE),
      skewness = ~EnvStats::skewness(., na.rm = TRUE),
      kurtosis= ~EnvStats::kurtosis(., na.rm = TRUE)
    ),
    .names = "{col}_{fn}")
    )

# present the results
summary_weight <- summary_weight |> 
  mutate(across(everything(), round, 2)) |>    # round to 3 decimal places
  pivot_longer(1:33, names_to = "Stats", values_to = "Values")  # long format

summary_weight|> 
  head(n = 11L)
```

## dlookr

```{r}
weight |>  
  dlookr::describe(dif_weight, birth_weight, discharge_weight) %>% 
  select(described_variables, n, na, mean, sd, p25, p50, p75, skewness, kurtosis) |> 
  ungroup() |> 
  print(width = 100)
```
:::
:::

As it was previously mentioned, the mean of the differences (-39.64) is close to median (-40). Moreover, both the skewness and the kurtosis are approximately zero indicating a symmetric and mesokurtic distribution for the weight differences.

**Normality test**

Additionally, we can check the statistical test for normality of the differences.

```{r}
 weight |> 
    shapiro_test(dif_weight)
```

The Shapiro-Wilk test suggests that the weight differences are normally distributed (p=0.74 \> 0.05).

## Run the paired t-test

We will perform a paired t-test to test the null hypothesis that the mean differences of weight equals to zero. Under this $H_o$, the test statistic is:

$$ 
t = \frac{\bar d}{SE_{\bar d}}
$$

where $\bar d$ is the mean of the differences, $SE_{\bar d} = s_d/ \sqrt n$ is the estimate of standard error and n is the number of pairs.

Under the null hypothesis, the t statistic follows the t-distribution with n-1 degrees of freedom (d.f.).

The 95% confidence interval of the mean of the differences is:

$$ 
95\% \ CI = \bar d \pm t_{n-1;a/2} * SE_{\bar d}
$$

**In R:**

First, we calculate the **mean of the differences**:

```{r}
mean_dif <- mean(weight$dif_weight, na.rm = TRUE)
mean_dif 
```

Second, we find the **standard error** of the mean of differences:

```{r}
n <- length(weight$dif_weight)
st_error <- sd(weight$dif_weight, na.rm = TRUE)/sqrt(n)
st_error
```

Therefore, the **t statistic** is:

```{r}
t <- mean_dif / st_error
t
```

The corresponding probability for this value can be calculated as the cumulative probability P(T \≤ t) for n-1 degrees of freedom. Then, the **p-value** is 2\*P(T \≤ t) because we are doing a two-tailed test.

```{r}
P <- pt(t, df = 24)

p_value <- 2*P
p_value
```

The **95% confidence interval** of the mean of differences is:

```{r}
lower_CI <- mean_dif + qt(0.025, df = 24)*st_error
lower_CI

upper_CI <- mean_dif - qt(0.025, df = 24)*st_error
upper_CI
```

Note that the 95% confidence interval (-53.0 to -26.3) doesn't include the null hypothesized value of zero.

 

Next, we present **R functions** to carry out all the tasks on our behalf.

::: callout-tip
## Paired t-test

::: panel-tabset
## Base R (1st way)

Our data are in a wide format. However, we are going to use only the `dif_weight` variable, inside the `t.test()`:

```{r}
t.test(weight$dif_weight)
```

## Base R (2nd way)

```{r}
t.test(weight$discharge_weight, weight$birth_weight, paired = T)
```

## rstatix

```{r}
weight |>  
  t_test(dif_weight ~ 1, detailed = T)
```
:::
:::

### Present the results

**Summary table**

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| 
tb1 <- weight |>  
  mutate(id = row_number()) |> 
  select(-dif_weight) |> 
  pivot_longer(!id, names_to = "group", values_to = "weights")

tb1 |>  
  tbl_summary(by = group, include = -id,
              label = list(weights ~ "weights (g)"),
              statistic =  weights ~ "{mean} ({sd})") %>%
  add_difference(test = weights ~ "paired.t.test", group = id,
                 estimate_fun = weights ~ function(x) style_sigfig(x, digits = 3)) |> 
  modify_table_body(~.x |> dplyr::relocate(stat_2, .before = stat_1)) |> 
  modify_table_body(~.x |> mutate(estimate = -estimate,
                                  ci = "-53.0, -26.3"))
```

**Report the results**

```{r}
t.test(weight$discharge_weight, weight$birth_weight, paired = T) |> 
  report()
```

  We can use the above information to write up a final report:

::: callout-tip
## Final report

There was a significant reduction of 39.6 g in weight after the discharge (mean change = -39.6 g, sd = 32.3[^paired_t_test-1], 95% CI: \[-52.96, -26.32\]; p-value \<0.001).
:::

[^paired_t_test-1]: sd for the change is useful information for meta-analytic techniques (see [Cochrane Handbook for Systematic Reviews of Interventions](https://training.cochrane.org/handbook/archive/v6.1/chapter-06))

We concluded that the result is statistical significant (reject $H_o$). However, is this reduction of clinical importance?
