# Probability Distributions {#sec-prob_distributions}

```{r}
#| include: false

library(fontawesome)
```

When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Understand and use the terminology of probability
:::

## Packages we need

We need to load the following packages:

```{r}
#| message: false

# graphs
library(ggplot2)
```

## Random Variables

A random variable assigns a numerical quantity to every possible outcome of a random phenomenon and may be:

-   discrete if it takes either a finite number or an infinite sequence of possible values
-   continuous if it takes any value in some interval on the real numbers

For example, a random variable representing the ABO blood system (A, B, AB, or O blood type) would be discrete, while a random variable representing the height of a person in centimeters would be continuous.

`r fa("arrow-right", fill = "orange")`   ***Example***

Suppose the X random variable for blood type is explicitly defined as follows:

$$X={\begin{cases}1, &for\ blood\ type\ A\\2, &for\ blood\ type\ B\\3, &for\ blood\ type\ AB\\4, &for\ blood\ type\ O\end{cases}}$$

That is, X is the discrete random variable that has four possible outcomes; it takes the value 1 if the person has blood type A, 2 if the person has blood type B, 3 if the person has blood type AB, and 4 if the person has blood type O.

We can also find the **probability distribution** that describes the probability of different possible values of random variable X. Note that the probability axioms and properties that we discussed earlier are also applied to random variables (e.g., the total probability for all possible values of a random variable equals to one).

Probability distributions are often depicted using probability tables or graphs. For example, assume that the individual probabilities for different blood types in a population are P(A) = 0.41, P(B) = 0.10, P(AB) = 0.04, and P(O) = 0.45 (Note that: P(A) + P(B) + P(AB) + P(O) = 0.41 + 0.10 + 0.04 + 0.45 = 1).

$$P(X=x)={\begin{cases}0.41,&for\ x=1\\0.10,&for\ x=2\\0.04,&for\ x=3\\0.45,&for\ x=4\end{cases}}$$

|                |      |      |      |      |
|----------------|:----:|:----:|:----:|:----:|
| **Blood type** |  A   |  B   |  AB  |  O   |
| **X**          |  1   |  2   |  3   |  4   |
| **P(X)**       | 0.41 | 0.10 | 0.04 | 0.45 |

Here, x denotes a specific value (i.e. 1, 2, 3, or 4) of the random variable X. Then, instead of saying P(A) = 0.41, i.e., the blood type is A with probability 0.41, we can say that P(X = 1) = 0.41, i.e., X is equal to 1 with probability of 0.41.

<br>

We can use the probability distribution to answer probability questions:

**What is the probability that a randomly selected person from the population can donate blood to someone with type B blood?**

We know that individuals with blood type B or O can donate to a person with blood type B. Therefore, we need to find the probability P(blood type B OR blood type O). Since the events blood type B or blood type O are mutually exclusive, we can use the addition rule for mutually exclusive events to get:

$$ \textrm{P(blood type B OR blood type O)= P(X = 2) + P(X = 4) = 0.10 + 0.45 = 0.55}$$

Hence, there is a 55% chance that a randomly selected person in our population can donate blood to someone with type B blood.

## Discrete probability distributions

The probability distribution of a discrete random variable X is defined by the **probability mass function** (pmf). Additionally, the **cumulative distribution function** (cdf) gives the probability that the random variable X is less than or equal to x and is usually denoted F(x).

 

### Bernoulli distribution

A random experiment with two possible outcomes, generally referred to as success (x = 1) and failure (x = 0), is called a **Bernoulli trial**.

Let X be a binary random variable of a Bernoulli trial which takes the value 1 (success) with probability p and 0 (failure) with probability 1-p. The distribution of the X variable is called Bernoulli distribution with parameter p, denoted as $X ∼ Bernoulli(p)$, where ${0\leq p\leq 1}$.

-   The probability mass function (pmf) of X is given by:

$$P(X=x)={\begin{cases}1-p,&for\ x=0\\p,&for\ x=1\end{cases}}$$ {#eq-bernoulli1}

which can also be written as: $$P(X=x)=p^{x}(1-p)^{1-x}\quad {\text{for }}x\in \{0,1\}$$ {#eq-bernoulli2}

 

-   The cumulative distribution function (cdf) of X is given by:

$$F(x) = P(X \le x)={\begin{cases}0,&for\ x <0\\1-p,&for\ 0\leq x < 1\\1,&for\ x \geq 1 \end{cases}}$$ {#eq-bernoulli3}

::: callout-note
The random variable X can take either value 0 or value 1. If $x<0$, then $P(X \le x) = 0$ because X can not take values smaller than 0. If $0\leq x < 1$, then $P(X \le x) = P(X=0) = 1-p$. Finally, if $x \geq 1$, then $P(X \le x) = P(X = 0) + P(X = 1) = (1 - p) + p = 1$.
:::

 

-   The mean of random variable, X, with Bernoulli(p) distribution is:

[**NOTE:** In this case, the mean can be interpreted as the proportion of the population who has the outcome (success).]{.aside}

$$μ = p$$ {#eq-bernouli4}

the variance is:

$$ \sigma^2= p(1-p)$$ {#eq-bernouli5}

and the standard deviation is:

$$σ = \sqrt{p(1-p)}$$ {#eq-bernouli6}

 

`r fa("arrow-right", fill = "orange")`   ***Example***

Let X be a random variable of the result of a surgical procedure, where X = 1 if the surgery was successful and X = 0 if it was unsuccessful. Suppose that the probability of success is p = 0.7. Then X has a Bernoulli distribution with parameter p = 0.7 :

$$X ∼ Bernoulli(0.7)$$

-   The pmf for this distribution is:

$$P(X=x)={\begin{cases}0.3,&for\ x=0\\0.7,&for\ x=1\end{cases}}$$ {#eq-bernouli7}

According to @eq-bernouli7 we have:

|          |     |     |
|----------|:---:|:---:|
| **X**    |  0  |  1  |
| **P(X)** | 0.3 | 0.7 |

 

We can plot the pmf for visualizing the distribution of the two outcomes (@fig-bernpdf).

```{r}
#| warning: false
#| label: fig-bernpdf
#| out-width: "80%"
#| fig-align: "center"
#| fig-cap: Plot of the pmf for Bernoulli(0.7) distribution.

x <- as.factor(c(0, 1))
y <- c(0.3, 0.7)
df1 <- data.frame(x, y)
 
# Plot
ggplot(df1, aes(x = x, y = y)) +
  geom_segment(aes(x = x, xend=x, y=0, yend = y), color = "black") +
  geom_point(color="deeppink", size = 4) +
  theme_classic(base_size = 14) +
  labs(title = "pmf Bernoulli(0.7)",
       x = "X", y = "Probability") +
  theme(axis.text = element_text(size = 14))
  
```

<br>

-   The cdf for this distribution is:

$$F(x) = P(X \le x)={\begin{cases}0,&for\ x <0\\0.3,&for\ 0\leq x < 1\\1,&for\ x \geq 1 \end{cases}}$$

```{r}
#| warning: false
#| label: fig-berncdf
#| out-width: "80%"
#| fig-align: "center"
#| fig-cap: Plot of the cdf for Bernoulli(0.7) distribution.

ggplot() +
  geom_segment(aes(x = -2, xend = 0, y = 0, yend = 0)) +
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 0.3), linetype = "dotted") +
  geom_segment(aes(x = 0, xend = 1, y = 0.3, yend = 0.3)) +
  geom_segment(aes(x = 1, xend = 1, y = 0.3, yend = 1), linetype = "dotted") +
  geom_segment(aes(x = 1, xend = 3, y = 1, yend = 1)) +
  theme_classic(base_size = 14) +
  labs(title = "cmf Bernoulli(0.7)", 
       x = "X", y = "F(x)") +
  theme(axis.text = element_text(size = 14))
```

 

### Binomial distribution

The binomial probability distribution can be used for modeling the number of times a particular event occurs (successes) in a sequence of *n* repeated and independent Bernoulli trials.

::: callout-note
# The binomial setting

1.  There is a fixed number of *n* repeated Bernoulli trials.
2.  The *n* trials are all independent. That is, knowing the result of one trial does not change the probability we assign to other trials.
3.  Both probability of success, *p*, and probability of failure, 1-p, are constant throughtout the trials.
:::

Let X be a random variable that indicates the number of successes in n-independent Bernoulli trials. If random variable X satisfies the binomial setting, it follows the binomial distribution with parameters *n* and *p*, denoted as $X ∼ Binomial(n, p)$, where *n* is the Bernoulli trial parameter (a positive integer) and *p* the Bernoulli probability parameter (${0\leq p\leq 1}$).

-   The probability mass function (pmf) of X is given by:

$$ P(X=x) = {{n}\choose{x}} \cdot p^x \cdot (1-p)^{n-x}$$ {#eq-binom0}

where x = 0, 1, ... , n, $\binom{n}{x} = \frac{n!}{x!(n-x)!}$, and ${0\leq p\leq 1}$.

Note that: $n! = 1\cdot 2 \cdot 3\cdot \ldots \cdot (n-2)\cdot (n-1)\cdot n$

 

-   The cumulative distribution function (cdf) of X is given by:

$$F(x) = P(X \le x)={\begin{cases}0,&for\ x <0\\\sum_{k=0}^{x}{\left( \begin{array}{c} n \\ k \end{array}
 \right) p^{k}(1 - p)^{n-k}},&for\ 0\leq x < n\\1,&for\ x \geq n \end{cases}}$$ {#eq-binom0.1}

 

-   The mean of random variable, X, with Binomial(n, p) distribution is:

$$μ = np$$ {#eq-binom1}

the variance is:

$$σ^2 = np(1-p)$$ {#eq-binom2}

and the standard deviation:

$$σ = \sqrt{np(1-p)}$$ {#eq-binom3}

 

`r fa("arrow-right", fill = "orange")`   ***Example***

Let the random variable X be the number of successful surgical procedures and suppose that a new surgery is successful 70% of the time (p = 0.7). If the results of 10 surgeries are randomly sampled, the X has a Binomial distribution $X ∼ Binomial(10, 0.7)$.

-   So, the pmf for this distribution is:

$$ P(X=x) = {{10}\choose{x}} \cdot 0.7^x \cdot (1-0.7)^{10-x}$$ {#eq-binom4}

The pmf of Binomial(10, 0.7) distribution specifies the probability of 0 through 10 successful surgical procedures.

According to @eq-binom4 we have:

|          |     |        |        |       |     |       |       |       |
|----------|:---:|:------:|:------:|:-----:|:---:|:-----:|:-----:|:-----:|
| **X**    |  0  |   1    |   2    |   3   | ... |   8   |   9   |  10   |
| **P(X)** |  0  | 0.0001 | 0.0014 | 0.009 | ... | 0.233 | 0.121 | 0.028 |

We can easily compute the above probabilities using the `dbinom()` function in R:

```{r}
binom_prob <- dbinom(0:10, size = 10, prob = 0.7)
binom_prob
```

We can plot the pmf for visualizing the distribution (@fig-binompmf).

```{r}
#| warning: false
#| label: fig-binompmf
#| out-width: "80%"
#| fig-align: "center"
#| fig-cap: Plot of the pmf for Binomial(10, 0.7) distribution.

df2 <- data.frame(x = 0:10, y = binom_prob)

# Plot
ggplot(df2, aes(x = x, y = y)) +
  geom_segment(aes(x = x, xend=x, y=0, yend = y), color = "black") +
  geom_point(color="deeppink", size = 4) +
  theme_classic(base_size = 14) +
  scale_x_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
  labs(title = "pmf Binomial(10, 0.7)",
       x = "X", y = "Probability") +
  theme(axis.text = element_text(size = 14))
```

<br>

-   The cdf for this distribution is:

$$F(x) = P(X \le x)={\begin{cases}0,&for\ x <0\\\sum_{k=0}^{x}{\left( \begin{array}{c} 10 \\ k \end{array}
 \right) 0.7^{k}(1 - 0.7)^{10-k}},&for\ 0\leq x < 10\\1,&for\ x \geq 10 \end{cases}}$$ {#eq-binom0.2}

In R, we can calculate the cumulative probabilities for all the possible outcomes using the `pbinom()` as follows:

```{r}
# find the cumulative probabilities
pbinom(0:10, size = 10, prob = 0.7)
```

The cdf for this distribution is shown below (@fig-binomcdf1):

::: panel-tabset
## geom_segment()

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
#| out-width: "80%"
#| label: fig-binomcdf1
#| fig-cap: Plot of the cdf for Binomial(10, 0.7) distribution.



# Plot
ggplot() +
  geom_segment(aes(x = -1, xend = 0, y = 0, yend = 0)) +
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 0.0000059049), linetype = "dotted") +
  
  geom_segment(aes(x = 0, xend = 1, y = 0.0000059049, yend = 0.0000059049)) +
  geom_segment(aes(x = 1, xend = 1, y = 0.0000059049, yend = 0.0001436859), linetype = "dotted") +
  
  geom_segment(aes(x = 1, xend = 2, y = 0.0001436859, yend = 0.0001436859)) +
  geom_segment(aes(x = 2, xend = 2, y = 0.0001436859, yend = 0.0015903864), linetype = "dotted") +

  geom_segment(aes(x = 2, xend = 3, y = 0.0015903864, yend = 0.0015903864)) +
  geom_segment(aes(x = 3, xend = 3, y = 0.0015903864, yend = 0.0105920784), linetype = "dotted") +

  geom_segment(aes(x = 3, xend = 4, y = 0.0105920784, yend = 0.0105920784)) +
  geom_segment(aes(x = 4, xend = 4, y = 0.0105920784, yend = 0.0473489874), linetype = "dotted") +
    
  geom_segment(aes(x = 4, xend = 5, y = 0.0473489874, yend = 0.0473489874)) +
  geom_segment(aes(x = 5, xend = 5, y = 0.0473489874, yend = 0.1502683326), linetype = "dotted") +
  
  geom_segment(aes(x = 5, xend = 6, y = 0.1502683326, yend = 0.1502683326)) +
  geom_segment(aes(x = 6, xend = 6, y = 0.1502683326, yend = 0.3503892816), linetype = "dotted") +
  
  geom_segment(aes(x = 6, xend = 7, y = 0.3503892816, yend = 0.3503892816)) +
  geom_segment(aes(x = 7, xend = 7, y = 0.3503892816, yend = 0.6172172136), linetype = "dotted") +
  
  geom_segment(aes(x = 7, xend = 8, y = 0.6172172136, yend = 0.6172172136)) +
  geom_segment(aes(x = 8, xend = 8, y = 0.6172172136, yend = 0.8506916541), linetype = "dotted") +
  
  geom_segment(aes(x = 8, xend = 9, y = 0.8506916541, yend = 0.8506916541)) +
  geom_segment(aes(x = 9, xend = 9, y = 0.8506916541, yend = 0.9717524751), linetype = "dotted") +
  
  geom_segment(aes(x = 9, xend = 10, y = 0.9717524751, yend = 0.9717524751)) +
  geom_segment(aes(x = 10, xend = 10, y = 0.9717524751, yend = 1), linetype = "dotted") +

  geom_segment(aes(x = 10, xend = 11, y = 1, yend = 1)) +
  theme_classic(base_size = 14) +
  scale_x_continuous(limits = c(-1, 11), breaks = seq(-1, 11, 1)) +
  labs(title = "cdf Binomial(10, 0.7)", x = "X", y = "F(x)") +
  theme(axis.text = element_text(size = 14))

```

## geo_step()

```{r}
#| warning: false
#| out-width: "80%"
#| fig-align: "center"

df3 <- data.frame(x = 0:10, y = pbinom(0:10, size = 10, prob = 0.7))

# Plot
ggplot(df3, aes(x=x, y=y)) +
  geom_step() +
  theme_classic(base_size = 14) +
  scale_x_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
  labs(title = "cdf Binomial(10, 0.7)",
       x = "X", y = "F(x)") +
  theme(axis.text = element_text(size = 14))
```
:::

<br>

**What is the probability that more than 8 of 10 surgical procedures are successful?**

We want to calculate the P(X \> 8):

$$ P(X > 8)= P(X = 9) + P(X = 10) = {{10}\choose{9}} \cdot 0.7^9 \cdot 0.3^1 + {{10}\choose{10}} \cdot 0.7^{10} \cdot 0.3^0 \Rightarrow $$

$$ P(X > 8)= 10 \cdot 0.04035 \cdot 0.3 + 1 \cdot 0.02824 = 0.12105 + 0.02825 = 0.1493$$

 

In R, we can calculate the probabilities P(X = 9) and P(X = 10) by applying the function `dbinom()` and adding the results:

```{r}
p9 <- dbinom(9, size=10, prob=0.7)
p9

p10 <- dbinom(10, size=10, prob=0.7)
p10

p9 + p10
```

 

Of note, another way to find the above probability is to calculate the 1-P(X ≤ 8):

```{r}
1 - pbinom(8, size=10, prob=0.7)
```

### Poisson distribution

While a random variable with a Binomial distribution describes a count variable (e.g., number of successful surgeries), its range is restricted to include integers from 0 to *n*. For example, the number of successful surgeries in a total of n = 10 surgical procedures cannot exceed 10.







