# R as calculator {#sec-rcalculator}

```{r setup_rcalculator, include=FALSE}

options(width=85)
knitr::opts_chunk$set(fig.pos = 'H')
```


R is a powerful tool for statistical computing and data analysis, but it can also serve as a basic calculator. In this chapter, we explore various arithmetic operations, simple comparisons with numbers, and the use of scientific notation (i.e., *e*-notation in R) for very small or large numbers. We also discuss how special values such as `NA`, `NaN`, and `Inf` behave in calculations.

 

## Arithmetic operators in R

One of the simplest tasks in R is performing basic arithmetic operations with numbers. For example:

```{r}

1 + 100 
```

Here, the `+` operator performs addition. The result is displayed with a preceding `[1]`, indicating the index of the output. @tbl-arithmeticoperators provides a list of the arithmetic operators available in R.


:::: {.columns}
::: {.column width="20%"}
:::
::: {.column width="60%"}


| Operator      | Description        |
|:---------------|:--------------------|
| \+            | addition           |
| \-            | subtraction        |
| \*            | multiplication     |
| /             | division           |
| \^ \ or \ **  | exponent           |
| %/%           | integer division   |
| %%            | modulo (remainder) |


: Basic arithmetic operators in R. {#tbl-arithmeticoperators .latex-table} 

:::
::: {.column width="20%"}
:::
::::

 

When using R as a calculator, it's important to note that the order of operations follows the same principles of arithmetic taught in high school mathematics, listed from highest to lowest precedence:

-   Parentheses: `( )`
-   Exponents: `^` or `**`
-   Multiplication, division, integer division, modulo: `*`, `/`, %/%, %% 
-   Addition, subtraction: `+`, `-`


For example:

```{r}
3 + 5 * 2
```

 

When multiple operators have the **same precedence**, they are evaluated in the order they appear in the mathematical expression from **left to right**, as shown below:

```{r}
 3 / 5 * 2   # the division 3/5 is done before the multiplication by 2
```

 

::: content-box-white
**INFO**

For more detailed information on how R prioritizes operators, we can refer to the R help section using the following command: 

**?Syntax**
:::

 


**Parentheses**

Parentheses are used to group operations, either to enforce a specific order of evaluation that differs from the default or to clarify the intended calculation.

```{r}
(3 + 5) * 2
```

 

Although it may sometimes seem redundant, using parentheses can clarify our intentions, which is particularly helpful when others review our code later.

```{r, eval=FALSE}

(3 + (5 * (2 ^ 2))) # hard to read
3 + 5 * 2 ^ 2       # clear, if we remember the rules
3 + 5 * (2 ^ 2)     # if we forget some rules, this might help
```



## Comparison operators in R

**Comparison (or relational) operators** are used to compare values. In R, comparisons return `TRUE` or `FALSE`, which can be abbreviated as `T` and `F` (though this is not recommended). Below is a list of comparison operators available in R (@tbl-relationaloperators).


:::: {.columns}
::: {.column width="20%"}
:::
::: {.column width="60%"}

| **Operator** | **Read as**              |
|:------------|:-------------------------|
| \<         | less than                |
| \>         | greater than             |
| ==         | equal to                 |
| \<=        | less than or equal to    |
| \>=        | greater than or equal to |
| !=         | not equal to             |

: Comparison (relational) operators in R. {#tbl-relationaloperators .latex-table} 

:::
::: {.column width="20%"}
:::
::::

 

Some simple comparisons with integer numbers follow:

```{r}
2 < 1  # less than
```

```{r}
1 > 0  # greater than
```

```{r}
1 == 1  # equal to (double equal sign for equality)
```

```{r}
1 <= 1  # less than or equal to
```

```{r}
-9 >= -3 # greater than or equal to
```

```{r}
1 != 2  # not equal to (inequality)
```


## Scientific notation and *e*-notation

Scientific notation is a method of expressing numbers that are very small or very large in a more compact and readable form. It typically follows the format $m \times 10^n$, where $m$ is a decimal number between 1 and 10, and $n$ is the exponent. In R, this is represented using **e**-notation (also called exponential notation), where **e** denotes exponentiation with base 10.

 

**Example (small number):** 0.000000015

-   **Scientific notation:** 0.000000015 = 1.5 × 0.00000001 = 1.5 × $10^{-8}$\
-   **e-notation:** 1.5e-08


```{r}
# type in Console and press Enter
0.000000015
```


 

**Example (large number):** 150000000

-   **Scientific notation:** 150000000 = 1.5 × 100000000 = 1.5 × $10^{8}$\
-   **e-notation:** 1.5e+08


```{r}
# type in Console and press Enter
150000000
```

 

To globally disable **e**-notation in R, we can set **`options(scipen = 999)`** at the beginning of our script. To restore the default behavior, simply reset `scipen` to 0, which will allow R to use **e**-notation when appropriate.



::: content-box-purple
**CAUTION!**

The **e**-notation in this context is not related to Euler's number ($e \approx 2.718$). 
:::

 


## Special values in R

There are a few special values that are used in R.

### Missing values (NA)

In the real world, missing data may occur when recording medical information (e.g., patients lost to follow-up, incomplete medical records). In R, the special numeric value `NA` stands for "Not Available" and represents a missing value. Arithmetic and comparison operations involving `NA` will return `NA`:

```{r}
1 + NA
```

```{r}
(3 + 5) / NA
```


```{r}
3 > NA
```

 

### Infinitive: -Inf or Inf

In R, there is a special value for infinity, denoted as `Inf`. This allows the representation of expressions such as:

```{r}
1 / 0
```

 

The value `Inf` can also be used in arithmetic operations, for example:

```{r}
Inf + 1000
```

 


### Not a Number (NaN)

The value `NaN` (stands for "Not a Number") represents an undefined or invalid value, typically resulting from certain arithmetic operations. For example:

```{r}
Inf / Inf
```

```{r}
0 / 0
```

```{r}
-Inf + Inf
```


 

### NULL

In R, `NULL`  is distinct from both `NA` and `NaN`. It represents the **null value** and is typically used in functions to indicate that an optional argument has not been assigned a value or that a function does not produce a meaningful result (see Chapter @sec-XX for more on functions). 
