# R functions {#sec-rfunctions}

```{r setup_rfunctions, include=FALSE}

options(width=85)
knitr::opts_chunk$set(fig.pos = 'H')
```


One of R's greatest strengths is its functional programming capabilities. In this chapter, we describe the main characteristics of functions in R, demonstrating them with a variety of examples. We explore mathematical functions, such as logarithmic and trigonometric functions, as well as other useful functions like **`round()`** and **`ifelse()`**.

 


## Characteristics of R Functions

Most computations in R involve the use of functions. An R function is a set of commands that are executed each time the function is called, performing a specific task. The standard version of R includes a number of built-in functions, and many others can be added or created by users (known as user-defined functions). We have already used some R functions to search for help, such as **`help()`** and **`example()`**.

In R, a function is called by typing its **name** followed by **parentheses**. The values provided inside the parentheses, separated by commas, are referred to as the **function's arguments**.


### Name and argumnets of a function

Let's have look at an example:

```{r}
seq(from = 5, to = 8, by = 0.5)
```


The function name is `seq`, and we pass three named arguments `from = 5`, `to = 8` and `by = 0.5`. The arguments `from = 5` and `to = 8` specify the starting and ending values of the sequence we want to create, while `by = 0.5` defines the increment between each value in the sequence.


The above result can also be obtained without naming the arguments, as shown below:

```{r}
seq(5, 8, 0.5)
```


::: content-box-white
**INFO**

We don’t need to specify the names of the arguments; instead, we can simply pass the values in the correct order, and R will match each value to its corresponding argument based on their position—this is known as positional matching.  
:::

 

Additionally, the **`seq()`** function provides several other arguments, which are documented on its help page and can be accessed by running **`?seq`**:

`seq(from = 1, to = 1, by = ((to - from)/(length.out - 1)), length.out = NULL, along.with = NULL, ...)`


For example, instead of specifying `by = 0.5`, we can use the argument `length.out = 26`  to create a sequence of 26 numbers as follows:

```{r}
seq(5, 8, length.out = 26)  # 26 numbers in the sequence
```


Note that when R prints a long sequence of values, it automatically splits the output across multiple lines for readability. The numbers in square brackets indicate the index of the first value on each line within the full sequence. In this example, `[1]` indicates the position of `5.00`, which appears first on the first line, while `[17]` indicates the position of `6.92`, which appears first on the second line. (*These indices may differ in your console, as the number of values displayed per line depends on the available console width.*)

 

Now, consider this command:

```{r}
seq(5, 8)
```

This generates a sequence running **from** 5 **to** 8. Since no step size is specified, the default increment step is calculated as **by**=((to - from)/(length.out - 1)) = (8-5)/(4-1) = 1, which is then passed to the `seq()` function.



### Required and optional arguments in functions

Let's explore another example. The **`log()`** is a mathematical function that calculates logarithms. To display its argument names along with their default values, we can use the **`args()`** function, as shown below:

```{r}
args(log)
```

In the **`log()`** function, `x` is a required argument, while `base` is an optional argument and has a default value of `exp(1)`.

-   **Required argument:** `x` is a required argument because if it is not provided to the function, the calculation will fail (i.e., the logarithm is undefined), resulting in an error:

```{r, error=TRUE, eval= FALSE}
log(base = 10)
```

`r kableExtra::text_spec("Error: argument x is missing, with no default", color = "#8676F8")`

 

-   **Optional argument:** If no value is provided for `base`, **`log()`** function will use `exp(1)` by default:

```{r}
log(15)      # in R, the default base for the log function is e (i.e., exp(1))
```


However, if we provide the argument `base = 10`, the **`log()`** function will calculate the base-10 logarithm:

```{r}
log(15, base = 10)      # the log function uses the specified base 10
```

 

::: content-box-purple
**IMPORTANT**

In R functions, some arguments are **required** and must be provided, while others are **optional** because they have default values specified in the function definition.
:::

 


Next, let's calculate the **natural logarithm of 3** (base = $e$):

```{r}
log(3)
```


All of the following expressions are equivalent:

```{r, eval=FALSE}
log(x=3)
log(x=3, exp(1))
log(x=3, base=exp(1))
log(3, exp(1))
log(3, base=exp(1))
log(base=exp(1), 3)
log(base=exp(1), x=3)
```


However, the following expression is not equivalent:

```{r}
log(exp(1), 3)
```


 

Moreover, it's important to note that not all functions have or require arguments to work. For example:

```{r}
date()
```

The **`date()`** function in R returns the current date and time as a character string.


## Mathematical functions

R has many built-in mathematical functions such as the `log(x)` that we have already seen.


### Logarithmic and Exponential functions

**_Logarithms_**

```{r}
log(5)     # natural logarithm (or base-e logarithm)
```

There are even built-in logarithmic functions with different bases:

```{r}
log2(5)    # base-2 logarithm
log10(5)   # base-10 logarithm
```

 

**_Exponents_**

```{r}
exp(5)     # Calculates Euler's number (e ≈ 2.718) raised to the power of 5
```

 

### Trigonometric functions

Trigonometric functions describe the associations between the sides and angles of a right-angled triangle. They also allow us to use angle measurements, either in radians or degrees, to determine the coordinates of any point on a circle, such as the unit circle. In R, trigonometric functions operate on angles measured in radians by default.

```{r}
sin(pi/2)    # sine of pi/2   (pi approximately equals to 3.14)
```


```{r}
cos(pi)      # cosine of pi
```


```{r}
tan(pi/3)    # tangent of pi/3
```

 

### Other mathematical functions

```{r}
sqrt(9)       # returns the squared root of a number
```


```{r}
abs(-9)       # returns absolute value of a number
```


```{r}
sign(-9)      # returns the sign of a number (-1 if negative, 0 if zero, 1 if positive)
```


```{r}
factorial(3)  # factorial (e.g., 3! = 1x2x3)
```



## Other useful functions


### The `round()` function

The **`round()`** function, which follows the **rounding principle**, is commonly used for rounding numbers. By default, it returns the nearest integer. For example:

```{r}
round(7 / 3)  # rounding 7/3 (2.333) to the nearest integer
```

To control the precision of the result, we can use the `digits` argument to specify the desired number of decimal places. For example:

```{r}
round(7 / 3, digits = 2)  # rounding 7/3 to two decimal places
```


When the first digit to be dropped is exactly 5, the **`round()`** function follows a  commonly used rule in programming: it rounds to the nearest **even** number. For example:

```{r}
round(1.5)  # it rounds 1.5 to the nearest even number, which is 2
round(2.5)  # it rounds 2.5 to the nearest even number, which is 2
```

 

There are a couple of additional functions that can be useful for rounding numbers:

```{r}
ceiling(16.2)       # round up to the nearest integer
```


```{r}
floor(16.2)         # round down to the nearest integer
```


```{r}
trunc(125.2395)     #  truncates the decimal places, removing any fractional part
```

 

### The `ifelse()` function

In R, the **`ifelse()`** function evaluates a logical condition and returns one of two specified values depending on whether the condition is `TRUE` or `FALSE`. The syntax is straightforward: `ifelse(condition, value_when_true, value_when_false)`. 


Let's demonstrate its usage with basic examples:

```{r}
ifelse(0.7 > 0.5, 1, 0)
```

In this example, **`ifelse()`** checks if 0.7 is greater than 0.5. Since the condition is `TRUE`, the function returns 1.


 

```{r}
ifelse(0.3 > 0.5, 1, 0)
```


Here, the condition is `FALSE` because 0.3 is not greater than 0.5, so the function returns 0.

 

## User-defined functions

We can create our own functions using **`function()`**, which provides a powerful way to extend R's capabilities and reuse code efficiently.


**The essential parts of a function are:**

1. The function’s name

2. The arguments (or parameters) of the function

3. The function's body (the code statements)

\vspace{10pt}

Here's a simple example of a function that calculates the logarithm of a number to base 7:

```{r}
log7 <- function(x) {
log(x, base = 7)
}

# calculate the logarithm of 5 to base 7
log7(5)
```


In this example, the function is defined by assigning the **`function(x)`** object to the name `log7` using the assignment operator (`<-`). (A detailed discussion of objects and assignment operators is presented in Chapter \@ref(robjects)). The `x` argument within the parentheses of **`function()`** serves as the input, while the code within the curly braces `{}` forms the function body—the code block that is executed when the function is called.

