# Lists and data frames {#sec-rlists}

```{r setup_rlists, include=FALSE}

options(width=85)
knitr::opts_chunk$set(fig.pos = 'H')
```


In this chapter, we explore the two types of generic vectors in R: lists and data frames. Lists offer a flexible way to store various data types, while data frames provide a structured format that is especially useful for working with tabular data.

All the required add-on packages for this chapter are part of the core set of packages within the **_tidyverse_** ecosystem. As such, we have the option to either load them collectively using the **_tidyverse_** meta-package or load them individually. For demonstration purposes, we choose the latter.

```{r, message=FALSE, warning=FALSE}

# Required packages for this chapter
library(tibble)             
library(dplyr)               
library(lubridate)           
```

 

## Creating a list

In R, a list allows us to store and organize a diverse range of objects, such as atomic vectors, matrices, and even other lists, within a single data structure. The objects within a list do not need to be associated or related to each other. Essentially, a list serves as an advanced data type, providing the flexibility to store nearly any type of information.

We can create a list using the **`list()`** function and assign it to an object named `my_list`, as shown below:

```{r}
my_list <- list(1:5, c("apple", "carrot"), c(TRUE, TRUE, FALSE))
my_list
```


The `my_list` consists of three elements, which we refer to as "list items" or simply "items". Specifically, these items are:

- A sequence of numbers from 1 to 5.

- A character vector containing the strings "apple" and "carrot".

- A logical vector with the values `TRUE`, `TRUE`, and `FALSE`. 

Therefore, each list item is an atomic vector of a different data type: numeric, character, and logical.

 

We can assign names to the list items as follows:

```{r}
my_list <- list(num = 1:5, 
                fruits = c("apple", "carrot"), 
                TF = c(TRUE, TRUE, FALSE))
my_list
```

Now, `my_list` is a named list containing three items: `num`, `fruits`, and `TF`.


We verify that the object’s class is `list`:
```{r}
class(my_list)
```

 

## Subsetting a list

### Subsetting a list and preserving output as a list

We can use the subsetting operator `[ ]` to select one or more list items while preserving the output in list format. Let's use the the previously created list as an example to demonstrate different methods of item selection.

- We can select the second list item by indexing its position:

```{r}

my_list[2]    

class(my_list[2])
```

The selected item is "fruits", and it is returned as a list.


- Alternatively, we can select the second list item by specifying its name within the square brackets:

```{r}

# same as above, but using the item's name
my_list["fruits"]   
```


- Finally, we can select the second list item using Boolean indices (`TRUE`/`FALSE`):

```{r}

# same as above, but using boolean indices 
my_list[c(FALSE, TRUE, FALSE)]    
```

Since `TRUE` is at the second position, the `[ ]` operator selects the second item in the list, and the output is preserved as a list.



### Subsetting a list and simplifying the output

We can use double square brackets `[[ ]]` to select one or more items from a list while simplifying the output:


- Select the second list item by indexing its position (simplifying the output):

```{r}

# select the second list item and simplify it to a vector
my_list[[2]]   

class(my_list[[2]])
```

When using double square brackets `[[ ]]`, R returns the actual data structure of the selected item, rather than preserving the list structure. Therefore, `my_list[[2]]` returns the character vector `c("apple", "carrot")`, not a list containing this vector.


- Similarly, specifying the name "fruits" within the double square brackets allows us to select this list item while simplifying the output:

```{r}
# same as above, but using the item's name
my_list[["fruits"]]   
```


- Another common method to select list items and simplify the output is by typing the name of the list (`my_list`), followed by a dollar sign (`$`), and then the name of the item within the list (e.g., "fruits").

```{r}

# select the list item "fruits" using the $ and simplify to a vector
my_list$fruits  
```

 

::: content-box-purple
**IMPORTANT**

-   One key difference between the [[ ]] operator and the \$ is that the [[ ]] can be used with both indices and names, while the \$ can only be used with names.

-   It is important to understand the distinction between simplifying and preserving when subsetting. Simplifying returns the simplest possible data structure that can represent the output, while preserving maintains the structure of the output the same as the input.
:::

 

### Selecting individual elements from a list item

By combining the `[[ ]]` or `$` operator with the `[ ]` operator, we can select individual elements from a specific list item. For example, to select the second element, "carrot", from the "fruits" item, we can use several approaches. 

- Select by indexing position:

```{r}
my_list[[2]][2] 
```



- Select by referencing the item name:

```{r}
my_list[["fruits"]][2]
```



- Select using `$` notation:
```{r}
my_list$fruits[2]
```

 

## Unlisting a list

To convert a list into an atomic vector, we can use the **`unlist()`** function:

```{r}
my_unlist <- unlist(my_list)
my_unlist
class(my_unlist)
```

 

## Recursive vectors and Nested Lists

In R, lists are sometimes referred to as **recursive vectors**\index{recursive vectors} because they can contain other lists within them. These inner lists are known as **nested lists**. For example:

```{r}
my_super_list <- list(item1 = 3.14,
                      item2 = list(item2a_num = 5:10,
                                   item2b_char = c("a", "b", "c")))

my_super_list
```

In this example, `item2` is a nested list, as it is the second item of `my_super_list`.

<br>

**Subsetting a nested list**

To select items from a nested list, we can combine the `[[ ]]` (or \$) operator and the `[ ]` operator.

- Select the `item2a_num` and preserve the output as a list:

```{r}
# preserve the output as a list
my_super_list[[2]][1]
class(my_super_list[[2]][1])
```



- Select the `item2a_num` and simplify it as a vector:

```{r}
my_super_list[[2]][[1]]
class(my_super_list[[2]][[1]])
```



- Select using names (simplify output):

```{r}
# same as above with names
my_super_list[["item2"]][["item2a_num"]]


# same as above with $ operator
my_super_list$item2$item2a_num

```

<br>

**Selecting individual elements from a nested list**

We can also **select individual elements** from the items of a nested list. For example, we select the "c" element as follows:

```{r}
# select individual element
my_super_list[[2]][[2]][3]
class(my_super_list[[2]][[2]][3])
```




Alternatively, we can use the `$` notation:

```{r}
my_super_list$item2$item2b_char[3]
```

 

## Data frames

A **data frame** is the most common way to store and organize rectangular data in R, and it is generally the preferred data structure for data analysis tasks. A data frame consists of **rows** and **columns**. While all elements within a column must have the same data type (e.g., numeric, character, or logical), different columns can have different data types. Therefore, a data frame is a special type of list with **equal-length** atomic vectors as its elements.

Various disciplines have different terms for the rows and columns in a data frame. In this textbook, we will consistently use the terms **"observations"** and **"variables"**, respectively.


### Creating a data frame

We will create a small example data frame with eight observations (rows) and six variables (columns) based on the following information:

-   **age:** age of the patient (in years). *Values = {30, 65, 35, 25, 45, 55, 40, 20}*.

-   **smoking:** smoking status of the patient (0=non-smoker, 1=smoker). *Values = {0, 1, 1, 0, 1, 0, 0, 1}*.

-   **ABO:** blood type of the patient based on the ABO blood group system (A, B, AB, O). *Values = {A, O, O, O, B, O, A, A}*.

-   **bmi:** Body Mass Index (BMI) category of the patient (1=underweight, 2=healthy weight, 3=overweight, 4=obesity). *Values = {2, 3, 2, 2, 4, 4, 3, 1}*.

-   **occupation:** occupation of the patient. *Values = {Journalist, Chef, Doctor, Teacher, Lawyer, Musician, Pharmacist, Nurse}*.

-   **adm\_date:** admission date to the hospital. *Values = {10-09-2023, 10-12-2023, 10-18-2023, 10-27-2023, 11-04-2023, 11-09-2023, 11-22-2023, 12-02-2023}*.



The data frame can be created using the **`data.frame()`** function in base R, the **`tibble()`** function in the **_tibble_** package, the [`data.table()`](https://rdatatable.gitlab.io/data.table/reference/data.table.html) function in the **_data.table_** package, or the [`tidytable()`](https://markfairbanks.github.io/tidytable/) function in the **_tidytable_** package. Let's try the **`tibble()`** :


```{r, message=FALSE, warning=FALSE}

dat <- tibble(age = c(30, 65, 35, 25, 45, 55, 40, 20),
              smoking = c(0, 1, 1, 0, 1, 0, 0, 1),
              ABO = c("A", "O", "O", "O", "B", "O", "A", "A"),
              bmi = c(2, 3, 2, 2, 4, 4, 3, 1),
              occupation = c("Journalist", "Chef", "Doctor", "Teacher",
                             "Lawyer", "Musician", "Pharmacist", "Nurse"),
              adm_date = c("10-09-2023", "10-12-2023", "10-18-2023", "10-27-2023",
                           "11-04-2023", "11-09-2023", "11-22-2023", "12-02-2023"))
dat
```



We can find the **type**, **class** and **dimensions** for the created object `dat`:

```{r, results='hold'}

typeof(dat)
class(dat)
dim(dat)
```

The type is a *list* but the class is a `tbl` *(tibble)* object which is a "tidy" data frame (tibbles work better in the tidyverse). The dimensions are 8x6.


### Accessing variables in a data frame

In R, we can access variables in a data frame just like items in a list by using their names or indices. For example:

```{r, results='hold'}

dat[["age"]]
dat[[2]]
```

or by using the **dollar sign (`$`)** :

```{r}
dat$age
```


We can also select individual elements of a specific variable as follows:

```{r}
dat$age[2:5]
```


### Converting columns to the appropriate data type

It is important to inspect the data types of columns and convert them to the appropriate type if necessary to ensure accurate analysis. The **`glimpse()`** function is commonly used to quickly examine the structure and data types of a data frame:

```{r}
glimpse(dat)
```


We observe a series of three-letter abbreviations in angle brackets (`<dbl>`, `<chr>`), which represent the data type of each column in a tibble. A list of these abbreviations is provided in @tbl-datatypes:

<br>

| **Type of Data**  | **Description**                                                                       | **Abbreviation** |
|:----------------:|:-----------------------------------|:----------------:|
|   character    | strings: letters, numbers, symbols, and spaces                                        |     `<chr>`      |
|    integer     | numerical values: integer numbers                                                     |     `<int>`      |
|     double     | numerical values: real numbers                                                        |     `<dbl>`      |
|    logical     | logical data, typically representing `TRUE` or `FALSE`                                |     `<lgl>`      |
|      date      | date (e.g, 2020-10-09)                                                                |     `<date>`     |
|   date+time    | date plus time (e.g., 2020-10-09 10:03:25 UTC)                                        |     `<dttm>`     |
|     factor     | categorical variables with fixed and known set of possible values (e.g., male/female) |     `<fct>`      |
| ordered factor | categorical variable with ordered fixed and known set of possible values              |     `<ord>`      |

: Tibble abbreviations that describe the type of data in columns of a data frame. {#tbl-datatypes .latex-table}

 

We can convert the categorical variables `smoking`, `ABO`, and `bmi` from `<dbl>`, `<chr>`, `<dbl>` types, respectively, to factors `<fct>` since they have fixed and known values.


-   **Variable: smoking** (numeric coded values → factor)

We will use the **`factor()`** function to convert the existing numeric coded (0/1) variable `smoking` to a factor\index{factor} with levels "non-smoker" and "smoker". 

```{r}
dat$smoking <- factor(dat$smoking, levels = c(0, 1), 
                      labels = c("non-smoker", "smoker"))
levels(dat$smoking)
```


<br>


-   **Variable: ABO** (chr → factor)

It's important to note that not all possible values may be present in a given dataset. For example, if we tabulate the variable `ABO` using the **`table()`** function, we obtain counts of the categories that are present in the data:

```{r}
# create a count table
table(dat$ABO)
```

The blood type "AB" of the ABO blood group system is absent from our data. In such cases, we can provide a list of all known levels in the **`factor()`** function:

```{r}

# create a vector containing the blood types A, B, AB, and O
ABO_levels <- c("A", "B", "AB", "O")

dat$ABO <- factor(dat$ABO, levels = ABO_levels)

# show the levels of status variable
levels(dat$ABO)
# create a count table
table(dat$ABO)
```

We observe that the count table includes the blood type "AB" even though it doesn't exist in the original data. 


<br>

-   **Variable: bmi** (numeric coded values → ordered factor)

We might have noticed that the categorical variable `bmi` takes numerically coded values (1, 2, 3, 4) in our dataset, so it is recognized as a double `<dbl>` type. We can convert this variable into an ordered factor `<ord>` with levels (1=underweight, 2=healthy, 3=overweight, 4=obesity). Instead of overwriting the existing variable, we prefer to create a new variable `bmi1`, as follows:"

```{r}
# create a vector containing the four bmi categories in order
bmi1_labels <- c("underweight", "healthy", "overweight", "obesity")

# create a new variable bmi1 as a factor with specified labels
dat$bmi1 <- factor(dat$bmi, levels = c(1, 2, 3, 4), 
                   labels = bmi1_labels, ordered = TRUE)
dat$bmi1
```

Now we can use, for example, the comparison operators `>` to check whether one element of the ordered vector is larger than the other.

```{r}
dat$bmi1[2] > dat$bmi1[6]
```

However, the use of these operators on factors is much less common compared to numeric vectors. Therefore, we typically omit the `ordered = TRUE` argument, especially when we provide the order of categories explicitly in the `levels` argument.

<br>

Suppose we want to categorize patients into broader BMI categories. For example, we can recode the original variable 'bmi' and create a new variable in our dataset called `bmi2`, merging "overweight" (coded as 3) and "obesity" (coded as 4) into a single category, such as "overweight".

```{r}
# create a new variable bm2 with recoded values
dat$bmi2 <- case_match(dat$bmi, 
                      1 ~ "underweight",
                      2 ~ "healthy",
                      c(3, 4) ~ "overweight")
```


In the code above, the **`case_match()`** function from the **_dplyr_** package matches the values on the left side of the tilde (~) against the input vector `dat$bmi`. For example, if the values 3 or 4 are found in the `bmi` column, they are recoded as "overweight" in the new variable.



Additionally, we can explicitly define the order of the levels for the `bmi2` variable:

```{r}
# set the levels in order
bmi2_levels <- c("underweight", "healthy", "overweight")

# convert bmi2 to a factor with specified labels
dat$bmi2 <- factor(dat$bmi2, levels = bmi2_levels)
dat$bmi2

dat
```


<br>

-   **Variable: adm_date** (chr → date)

The standard date format in R is `YYYY-MM-DD`. If the original data represents dates in a different format, we need to convert them to the standard format. 

The **_lubridate_** package provides functions for formatting dates by specifying the order of day (`d`), month (`m`), and year (`y`). For example, if a dataset contains dates in the format `MM-DD-YYYY` (i.e., month-day-year), the **`mdy()`** function converts them into date objects. Likewise, if the dates are in the format `DD-MM-YYYY` (i.e., day-month-year), the **`dmy()`** function performs the conversion. In our example, the dates are in the `MM-DD-YYYY` format, so we must convert them accordingly.

```{r, message=FALSE, warning=FALSE}

dat$adm_date <- mdy(dat$adm_date)
dat
class(dat$adm_date)
```


In the code above, the **`mdy()`** function from the **_lubridate_** package parses dates in the "month-day-year" format and convert them into R's standard date format (`YYYY-MM-DD`).

