# R objects {#sec-robjects}

```{r setup_robjects, include=FALSE}

options(width=85)
knitr::opts_chunk$set(fig.pos = 'H')
```

In this chapter, we introduce the concept of objects in R, which is fundamental to working effectively with the language. Objects can take many forms, and R manages all of them, both pre-defined and user-created, in a central workspace called the global environment. Assigning valid and meaningful names to objects using the assignment operator (`<-`) is essential for writing clear, organized, and shareable code.

 

## What Are the Objects in R?

R works with **objects**, which include everything we interact with or encounter in R, including numbers, data structures, functions, and outputs such as plots. Objects can either come from R packages or be created by the user. User-created objects are assigned names specified by the user. R stores these objects in the global environment, enabling easy access and manipulation throughout the R session.

In R, objects typically have several properties, known as **attributes**, which define their structure and guide how R interprets them. We can access an object’s attributes using the **`attributes()`** function. However, not all R objects have attributes; in such cases, **`attributes()`** returns `NULL`.

For example, the well-known `anscombe` dataset, which contains 11 rows and 8 columns, has three key attributes. These include `names` (the column names), `class` (which is `data.frame`), and `row.names` (which represent the row identifiers).

```{r}
attributes(anscombe)
```

 

Two relative functions that reveal a dataset’s structure are **`class()`** and **`dim()`**:

```{r}
class(anscombe)
dim(anscombe)   
```


Note that the **`dim()`** function is particularly useful because dimensions are not stored as an explicit attribute; instead, they are derived from the lengths of `row.names` and `names`.

## Named Storage of Objects

### Assignment Operator (`<-`)

In R, the recommended way to assign a value to an object is by using the left-arrow assignment operator (`<-`), which combines the less-than sign (`<`) and the hyphen (`-`) (keyboard shortcut: `r kableExtra::text_spec("Alt", color = "#8676F8", bold = TRUE)` + `r kableExtra::text_spec("-", color = "#8676F8", bold = TRUE)` for Windows/Linux and `r kableExtra::text_spec("Option", color = "#8676F8", bold = TRUE)` + `r kableExtra::text_spec("-", color = "#8676F8", bold = TRUE)` for Mac).

For example, to store the value `1/40` in an object named `x`, we assign it as follows:

```{r}
x <- 1/40       # assignment (right to left)
```

Note that assignment does not automatically print the result to the Console. Instead, R stores the value in the **object** `x` for later use. If we check the `Environment` tab in RStudio, we will see `x` and its corresponding value listed there. To retrieve the stored value, we can simply call the object `x`, as shown below:

```{r}
x
```

However, if the assignment expression is enclosed in parentheses, R will evaluate the expression and display the result immediately. For example:

```{r}
(x <- 1/40)
```

 

It is important to include **spaces before and after** comparison operators and assignment operators. For example, consider evaluating the expression "x less than -1/50" (note that the current value of `x` is `1/40`):

-   **With** spaces

```{r}
x < -1/50    # with spaces 
```

The result is `FALSE` because the value of `x` (which is 1/40) is greater than -1/50.

 

-   **Without** spaces

```{r}
x<-1/50    # without spaces
x
```

If we omit the spaces, the left assignment operator is used instead of the comparison operator, leading to `x <- 1/50`, which equals 0.02. As a result, the value of `x`, which was originally 1/40, is reassigned to 1/50.

 

It is important to note that the object `x` can also be used in place of a numeric value in any function that expects a numeric input. For example:

```{r}
log(x)     # given that x is currently 1/50, this computes log(1/50)
```

Additionally, assignment expressions can include the object being assigned to, as demonstrated below:

```{r}

x <- x + 1  
x
```

In this expression, `x` on the right-hand side refers to the current value of `x` (1/50), and the expression adds 1 to it **before** assigning the result back to `x`.

 

### Other Types of Assignment Operators

It is also possible to use the equal sign (`=`) for assignment or the right-arrow assignment operator (`->`) for rightwards assignment, though these are less commonly preferred by R users.

For example:

```{r}
x = 1/40      # assignment (right to left), equivalent to <-
x
```

or

```{r}
1/40 -> x     # assignment (left to right)
x
```

It's advisable to **be consistent** with the assignment operator we use.

## Valid Object Names

Valid object names must adhere to specific rules: they must start with a letter and **may** include letters, numbers, underscores ( `_` ), and periods (`.`). However, they **cannot** contain spaces, or use [Reserved words](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html) such as `TRUE`, `FALSE`, or `NA`, as these have special meanings in R. When naming an object with multiple words, various naming conventions can be used. Common styles include periods.between.words, underscores_between_words, and camelCaseToSeparateWords.

The choice of what we use is up to us, but it's crucial to maintain consistency. If assistance is needed, we can refer to:

```{r, eval=FALSE}

??make.names
??clean_names
```

It is also important to note that R is case-sensitive, meaning it distinguishes between uppercase and lowercase letters.

```{r}
Y <- 50
Y
```

If we try to access the lowercase y instead, R will return an error because it does not recognize it as the same object:

```{r, eval=FALSE}
y
```

`r kableExtra::text_spec("Error: object 'y' not found", color = "#8676F8")`

## Diverse Storage in Objects

Objects can store more than just numbers, including strings of characters, which must be enclosed in quotes. For example:

```{r}
sentence <- "the Fellowship of the Ring consisted of Frodo plus 8 others"
```

The data type stored in an object determines which operations can be performed on it. For instance, if a number is stored as a character string instead of a numeric type, arithmetic operations won’t work. Consider this example:

```{r, eval=FALSE}
eight <- "8"
eight + 1
```

Since eight is a character string here, R cannot perform addition, resulting in this error:

`r kableExtra::text_spec("Error in eight + 1: non-numeric argument to binary operator", color = "#8676F8")`


To apply arithmetic operations, the object must contain a numeric value. We can convert the character "8" to numeric with `as.numeric()` function before adding (see @sec-rvectors) for the concept of "explicit coercion").

```{r}
eight <- "8"
as.numeric(eight) + 1
```

 


Notice that R commands are typically written on separate lines, but they can also be combined on a single line by separating them with a semicolon (`;`). For example, the previous two commands could be written on one line like this:

```{r}
eight <- "8"; as.numeric(eight) + 1
```

While this approach is syntactically valid, it is generally discouraged because it reduces code readability and makes debugging more challenging.
