# Receiver Operating Characteristic (ROC) curve {#sec-roc}

```{r}
#| include: false

library(fontawesome)
library(htmlwidgets)
```


When a diagnostic test result is measured in a **continuous scale**, sensitivity and specificity vary with different cut-off points (thresholds). Therefore, a **convenient** cut-off point[^roc-1] must be selected in order to calculate the measures of diagnostic accuracy of the test. **Receiver operating characteristic** (ROC) curve analysis can be used to help with this decision.

[^roc-1]: **NOTE:** there is critique, however, that the binary diagnosis is problematic.

When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Evaluate a diagnostic test
-   Compare different diagnostic tests
-   Interpret the results
:::

## Research question

We want to compare two screening questionnaires for chronic obstructive pulmonary disease (COPD) among smokers aged \>45 years in the primary care setting:

-   *International Primary Care Airways Group (IPAG)* questionnaire (Score: 0-38)
-   *COPD Population Screener (COPDPS)* questionnaire (Score: 0-10)

Each participant received both questionnaires ('fully paired' design). The diagnosis of COPD was based on spirometric criterion (FEV 1 /FVC \<0.7 following bronchodilation), clinical status (medical history, symptoms and physical examination), and exclusion of other diseases.

## Packages we need

We need to load the following packages:

```{r}
#| message: false
#| warning: false

library(pROC)
library(plotROC)
library(epiR)
library(ggsci)
library(here)
library(tidyverse)
```

> **Other relative packages:** OptimalCutpoints, cutpointr

## Preparing the data

We import the data *copd* in R:

```{r}
#| warning: false

library(readxl)
dat <- read_excel(here("data", "copd.xlsx"))
```

```{r}
#| echo: false
#| label: fig-depression
#| fig-cap: Table with data from "copd" file.

DT::datatable(
  dat, extensions = 'Buttons', options = list(
    dom = 'tip',
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)

```

We inspect the data and the type of variables:

```{r}
glimpse(dat)
```

## Using cut-off points and the ROC curve

Based on previous studies, the cut-off points for a positive response are:

-   ≥ 17 for the IPAG questionnaire
-   ≥ 5 for the COPDPS questionnaire.

We can evaluate these cut-off values by calculating their associated measures of diagnostic accuracy (i.e Se, Sp, PPV, NPV).

```{r}
#| warning: false

dat <- dat |>  
  mutate(IPAG_cat = cut(IPAG, c(0, 17, 38), labels=c("-","+"), 
                        include.lowest = TRUE, right=FALSE),
         COPDPS_cat = cut(COPDPS, c(0, 5, 10), labels=c("-","+"), 
                          include.lowest = TRUE, right=FALSE))

dat <- as.data.frame(dat)

# we need to create a roc object for each questionnaire
roc1 <- roc(dat$diagnosis, dat$IPAG)
roc2 <- roc(dat$diagnosis, dat$COPDPS)
```

::: callout-important
## NOTE

-   For screening purposes such as mammogram, the cut-off point can be selected to favor a higher sensitivity. Thus, a negative test result indicates the absent of the disease (Se-N-out; sensitive, negative, "rule out" the disease).

-   For confirmative diagnosis purposes, for example, when a chemotherapy is to initiated once the diagnosis is established, the cut-off point can be selected to favor a higher specificity. Thus, a positive test result indicates the presence of disease (Sp-P-in; specificity, positive, "rule in" the disease).
:::

Additionally, for a given diagnostic test, we can consider all cut-off points that give a unique pair of values for sensitivity and specificity. We can plot in a graph, which is known as a **ROC curve**, the sensitivity on the y-axis and 1-specificity (false positives) values on the x-axis for all these possible cut-off points of the diagnostic test. Then, the **area under the ROC curve** (AUC of ROC), also called the c-statistic, can be calculated which is widely used as a measure of overall performance.

### IPAG questionnaire

**A. The use of a cut-off value: IPAG score ≥17**

First, we will find the counts of individuals in each of the four possible outcomes in a 2×2 table for the cut-off point of 17:

```{r}
table(dat$IPAG_cat, dat$diagnosis)
```

Next, we reformat the table as follows:

```{r}
tb1 <- as.table(
  rbind(c(203, 644), c(70, 1670))
  )

dimnames(tb1) <- list(
  Test = c("+", "_"),
  Outcome = c("+", "-")
)

tb1
```

```{r}
epi.tests(tb1, digits = 3)
```

The results using the cut-off point of 17 give Se = 0.744 (0.687 - 0.794) and Sp = 0.722 (0.703 - 0.740). We observe that the probability of the absence of COPD given a negative test result is high NPV = 0.960 (95% CI: 0.949, 0.969) in this sample with smokers.

 

**B. The area under the ROC curve of IPAG questionnaire**

Let's calculate the AUC of ROC for the IPAG questionnaire:

```{r}
#| warning: false
auc(roc1)
```

The 95% confidence interval of this area is:

```{r}
#| warning: false

ci.auc(roc1)
```

The ability of the IPAG questionnaire to discriminate between individuals with and without COPD is shown graphically by the ROC curve in (@fig-roc1):

```{r}
#| warning: false
#| label: fig-roc1
#| fig-width: 7
#| fig-cap: The ROC curve of IPAG questionnaire.


# create the plot
g1 <- ggplot(dat, aes(d = diagnosis, m = IPAG)) + 
  geom_roc(n.cuts = 0, color = "#0071BF") +
  theme(text = element_text(size = 14)) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  scale_x_continuous(expand = c(0, 0.015)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "1 - Specificity", y = "Sensitivity")

# add annotations to the plot
g1 + annotate("text", x=0.70, y=0.30, 
           label=paste("AUC IPAG = ", 0.799, 
                       "(95% CI = ", 0.769, " - ", 0.829, ")"))
```

The AUC of IPAG questionnaire equals to 0.799 (95% CI: 0.769 - 0.829) which indicates a reasonable[^roc-2] diagnostic test.

[^roc-2]: The dashed **diagonal** line connecting (0,0) to (1,1) in the ROC plot corresponds to a test that is completely useless in diagnosis of a disease, AUC = 0.5 (i.e. individuals with and without the disease have equal "chances" of testing positive). A test which is **perfect** at discriminating between those with disease and those without disease has an AUC = 1 (i.e. the ROC curve approaches the upper left-hand corner).

::: callout-note
**Youden index** (*J* statistic), which is defined as the sum of sensitivity and specificity minus 1, is often used in conjunction with the ROC curve. The maximum value of the *Youden index* may be used as a criterion for selecting the optimal cut-off point (threshold) for a diagnostic test as follows:

```{r}
#| warning: false

coords(roc1, "best", ret = c("threshold", "sensitivity", "specificity"),
       best.method="youden")
```

We observe that the optimal cut-off point (threshold) for this sample equals to 18.5 which is slightly higher than the value of 17 that was obtained from other studies.

We can also easily calculate the maximum value of *Youden index* according to the previous definition of the Youden index:

```{r}
0.674 + 0.806	- 1
```
:::

 

### COPDPS questionnaire

**A. The use of a cut-off value: COPDPS score ≥5**

First, we will find the counts of individuals in each of the four possible outcomes in a 2×2 table for the cut-off point of 5:

```{r}
table(dat$COPDPS_cat, dat$diagnosis)
```

Next, we reformat the table as follows:

```{r}
tb2 <- as.table(
  rbind(c(152, 225), c(121, 2089))
  )

dimnames(tb2) <- list(
  Test = c("+", "_"),
  Outcome = c("+", "-")
)

tb2
```

```{r}
epi.tests(tb2, digits = 3)
```

The results using the cut-off point of 5 give Se = 0.577 (0.496 - 0.617) and Sp = 0.903 (0.890 - 0.915).

 

**B. The area under the ROC curve of COPDPS questionnaire**

The AUC of ROC curve of COPDPS questionnaire is:

```{r}
#| warning: false

auc(roc2)
```

The 95% confidence interval of this area is:

```{r}
#| warning: false

ci.auc(roc2)
```

The ROC curve of COPDPS questionnaire (@fig-roc2) follows:

```{r}
#| warning: false
#| label: fig-roc2
#| fig-width: 7
#| fig-cap: The ROC curve of COPDPS questionnaire.


# create the plot
g2 <- ggplot(dat, aes(d = diagnosis, m = COPDPS)) + 
  geom_roc(n.cuts = 0, color = "#EFC000") +
  theme(text = element_text(size = 14)) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  scale_x_continuous(expand = c(0, 0.015)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "1 - Specificity", y = "Sensitivity")

# add annotations to the plot
g2 + annotate("text", x=0.70, y=0.25, 
           label= paste("AUC COPDPS = ", 0.791, 
                        "(95% CI = ", 0.760, " - ", 0.821, ")"))
```

The AUC of COPDPS questionnaire equals to 0.791 (95% CI: 0.760 - 0.821) which is close to the value 0.799 of AUC of IPAG questionnaire.

::: callout-note
The optimal cut-off point using the *Youden Index* as best method is:

```{r}
#| warning: false

coords(roc2, "best", ret = c("threshold", "sensitivity", "specificity"),
       best.method="youden")
```

We observe that the optimal cut-off point (threshold) for this sample equals to 4.5 which is close to the value of 5 that was obtained from other studies.

We can also calculate the maximum value of the *Youden index*:

```{r}
0.557 + 0.903	- 1
```
:::

## Comparing ROC Curves

**A. Graphical comparison of ROC curves**

We can plot the ROC curves for both questionnaires in the same graph and compare the area under the curves (@fig-rocs):

```{r}
#| warning: false
#| label: fig-rocs
#| fig-width: 7
#| fig-cap: Graphical comparison between IPAG and COPDPS ROC curves.

# prepare the data
longdata <- melt_roc(dat, "diagnosis", c("IPAG", "COPDPS"))

# create the plot
g <- ggplot(longdata, aes(d = D, m = M, color = name)) + 
  geom_roc(n.cuts = 0) +
  theme(text = element_text(size = 14),
        legend.position="top") +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  scale_x_continuous(expand = c(0, 0.015)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_color_jco() +
  labs(x = "1 - Specificity", y = "Sensitivity", colour="Questionnaire")

# add annotations to the plot
g + annotate("text", x=0.70, y=0.35, color = "#0071BF",
           label=paste("AUC IPAG = ", 0.799, 
                       " (95% CI = ", 0.769, " - ", 0.829, ")")) +
  annotate("text", x=0.70, y=0.28, color = "#EFC000",
           label= paste("AUC COPDPS = ", 0.791, 
                        "(95% CI = ", 0.760, " - ", 0.821, ")"))
```

The AUC values obtained from the ROC curve were 0.799 (95% CI: 0.769 - 0.829) for the IPAG questionnaire and 0.791 (95% CI: 0.760 - 0.821) for the COPDPS questionnaire. Therefore, the two questionnaires have similar overall performance in the present sample.

 

**B. Compare AUCs using the DeLong' s test**

The DeLong's test can be used for comparing 2 areas under the curve (AUCs).

::: {.callout-note icon="false"}
## Null hypothesis and alternative hypothesis

-   $H_0$: there is no difference between two AUCs ($AUC_{IPAG} = AUC_{COPDPS}$)
-   $H_1$: there is difference between two AUCs ($AUC_{IPAG} \neq AUC_{COPDPS}$)
:::

```{r}
#| warning: false

roc.test(roc1, roc2, method=c("delong"))

```

There was no significant difference in the AUC values with the two questionnaires (p = 0.45 \< 0.05).
