# Survival analysis {#sec-survival}

In health sciences we often have data which represent the **time until some event occurs**. In a clinical trial setting, the primary outcome may be the time from onset of therapy until a well-defined critical event such as death, cancer recurrence, or first occurrence of a particular adverse event. One of the challenges specific to survival data is that the survival times will be unknown for some patients (censored data) for a variety of reasons:

-   lost to follow-up during the study period
-   withdrawn from the study because of some reason
-   the event did not occur by the end of the study

Such data could be referred to as **right censored** because, on a timeline, the true lifetimes of the patients are to the right of their observed censor times.

Statistical techniques for survival data have been developed that are be able to account for censored observations. In survival analysis, it is generally assumed that censoring is **non-informative** of survival (being censored or not is not related to the probability of the event occurring).

Survival analysis is a class of statistical methods. One approach to estimate the the survival over time is the **Kaplan-Meier** procedure which is a non-parametric method (i.e. it does not make any assumptions about the underlying distribution of the data).

When we have finished this Chapter, we should be able to:

::: {.callout-caution icon="false"}
## Learning objectives

-   Understand the basic concepts of the Kaplan-Meier analysis of time-to-event data
-   Compare different survival curves
-   Interpret the results
:::

## Research question

In a randomized controlled trial designed to find out the efficacy of a new therapy for leukemia, the patients were randomly assigned into two groups (new therapy versus standard therapy). The researchers want to (i) estimate the survival time of patients receiving the new therapy, and (ii) compare the survival curves of the patients receiving the new therapy and patients receiving the standard therapy.

## Packages we need

We need to load the following packages:

```{r}
#| message: false
#| warning: false

# packages for graphs
library(ggplot2)
library(survival)
library(ggsurvfit)
library(ggprism)
#library(survminer)
library(asaur)
library(knitr)
library(kableExtra)
library(gt)
library(gtExtras)

library(here)
library(tidyverse)
```

## Preraring the data

In a clinical trial, the patients are monitored for the occurrence of a particular event or outcome. In preparing Kaplan-Meier survival analysis that compares different treatments, three variables should be recorded for each patient: (a) the **time** that is the duration between the beginning of the treatment and the end-point (event of interest or censoring), (b) the **status** at the end of the survival time (event occurrence or censored data), and (c) the **study group** such as treatment versus control intervention.

We import the data *"leukemia"* in R:

```{r}
#| warning: false

library(readxl)
dat <- read_excel(here("data", "leukemia.xlsx"))
```

```{r}
#| echo: false
#| label: fig-depression
#| fig-cap: Table with data from "leukemia" file.

DT::datatable(
  dat, extensions = 'Buttons', options = list(
    dom = 'tip',
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  )
)
```

The dataset has 42 observations (rows) and includes three variables (columns).

-   **time:** the survival or censoring time in months

-   **status:** indicator whether or not the patient died (1 indicates death and 0 indicates censored observation)

-   **intervention:** randomly assigned therapy group with two levels, therapy A (new) or therapy B (standard).

We inspect the data and the type of variables:

```{r}
glimpse(dat)
```

First, we obtain a subset of data with patients receiving the new therapy A :

```{r}
dat.A <- dat |> 
  filter(intervention == "therapy A (new)")
```

Next, we run the **`Surv()`** function that converts the data to a special format which allows to account for censored observations (i.e. a compiled version of the survival time and status). The arguments we need to provide are the variables *time* and *status* of the patients. The times (in months) for the patients receiving the new therapy A follow:

```{r}
Surv(dat.A$time, dat.A$status)
```

**NOTE:** Censoring times are marked with "**+**" symbol. The above data show that patients in the new therapy were censored on months 6+, 9+, 10+, 11+, 17+, 19+, 20+, 32+, 32+, 34+, 35+.

## The Kaplan--Meier Product Limit Estimator

Kaplan-Meier estimator, also known as product-limit estimator, can be used to measure the cumulative probability of "surviving" for a certain amount of time after starting the therapy.

### Basic concepts

Let $T$ be a non-negative random variable, representing the time until the event of interest (death). Additionally, suppose that the events (deaths) are observed in the period of follow-up at $k$ distinct times $t_{1} < t_{2} < t_{3} < \ ...\ < t_{k}$. The **conditional survival probability** at $t_j$ is defined as the probability of surviving beyond time $t_j$ ($T>t_j$), given that the patient has survived at least $t_j$ time ($T≥t_j$). This conditional probability represents the proportion of patients who are at risk at time $t_j$ but who do not die at this time point, as follows:

$$P(T > t_j\ |\ T≥t_{j})  = \frac{r_{j}-d_{j}}{r_{j}}\ = 1-\frac{d_{j}}{r_{j}} , \ \ \ for\ j=1,2,..,k. $$ {#eq-cond_prob}

where

$t_{j}$ is the time when at least one event (i.e., death) happened

$r_j$ is the number of patients at risk (i.e., the number of patients alive) just before the time $t_j$

$d_j$ the number of events (i.e., deaths) that happened at time $t_j$

 

The **Kaplan--Meier estimator** for the survival function $S(t)$ is defined as the **cumulative product** of the conditional survival probabilities:

$$ S(t) = P(T>t)  =  \prod_{j:t_j≤t}(1-\frac{d_{j}}{r_{j}})$$ {#eq-cum_surv1}

Therefore, the cumulative probability of surviving beyond $t_j$ is given by:

$$ S(t_{j}) = (1-\frac{d_{j}}{r_{j}}) \times (1-\frac{d_{j-1}}{r_{j-1}})   \times \ ... \times\ (1-\frac{d_{2}}{r_{2}}) \times (1-\frac{d_{1}}{r_{1}}) $$ {#eq-cum_surv2}

This implies that:

$$ S(t_{j}) =  (1-\frac{d_{j}}{r_{j}}) \times S(t_{j-1}) $$ {#eq-cum_surv3}

### K-M analysis for the group of patients in new therapy A

The first step in Kaplan-Meier analysis usually involves the construction of a table with the Kaplan--Meier estimates, so we call the `survfit2()` function that models the survival probability with the formula `Surv ∼ 1` (we use 1 as we have not any grouping variable):

```{r}
#| message: false
#| warning: false

# create an object with the K-M estimates for the patients in the new therapy A
km.A <- survfit2(Surv(time, status) ~ 1, data = dat.A)

# obtain the list of variables included in the Km.A object
names(km.A)
```

As we can see, the function `survfit()` returns a list of variables including the number of participants at risk (`n.risk`), censored (`n.censor`) and having experienced an event (`n.event`) as well as the cumulative probability of surviving over time (`surv`) for the new therapy, among others.

Next, we estimate in R the conditional probability of surviving over time using the @eq-cond_prob:

```{r}
#| message: false
#| warning: false

# compute the probability of "surviving" over time
prob.A <- round(1-(km.A$n.event/km.A$n.risk), 3)
prob.A
```

 

Now, we are ready to include all the information in one table (@tbl-tb1) as follows:

```{r}
#| message: false
#| warning: false
#| label: tbl-tb1
#| tbl-cap: "sdf sfsf"

# create a dataframe with all the "survival" variables of interest
tb.A <- data.frame(time = km.A$time, n.risk = km.A$n.risk,
                   n.event = km.A$n.event, n.censor = km.A$n.censor,
                   prob.A, surv.A = round(km.A$surv, 3))

# create the table
tb.A |> 
  gt() |> 
  cols_label(
  time = html("Ordered times (months), t"),
  n.risk = html("No. at risk"),
  n.event = html("No. of deaths"),  
  n.censor = html("No. censored"),
  prob.A = html("Conditional probability of surviving, P(t)"),
  surv.A = html("Cumulative probability of surviving, S(t)")) |>
  tab_options(column_labels.font.weight = "bold") |> 
  cols_align(align = "center") |> 
  gt_highlight_rows(rows = 3, fill = "lightgrey",
                    bold_target_only = TRUE, 
                    target_col = c(n.risk, n.censor)) |> 
  tab_style(style = list(cell_text(weight = "bold")),
            locations = cells_body(columns = n.risk, rows = n.risk == 15))
```

When there are only censored observations at a particular time such as at month 9, the conditional probability $P(t)$ of surviving equals to 1 and the cumulative probability $S(t)$ does not changed, $S(9) = S(7) = 0.807$. However, we observe that at the next time t = 10 months, the number of patients "at risk" is reduced by the number of censored data at t = 9 months ($n.risk = 16 - 1 = 15$). The conditional probability for this time point equals to $P(10) = 1-(1/15) = 1-0.067 = 0.933$ because one patient died. Thus, the cumulative probability of surviving beyond 10 months becomes $S(10) = P(10) \times S(9) = 0.933 \times 0.807 = 0.753$.

::: callout-important
Only events cause the survival curve to drop. Censoring causes a larger drop for the next event because it reduces the number of patients at risk when that next event occurs.
:::

### The Kaplan--Meier curve

The cumulative survival probability is calculated at each time $t_j$ and represents the proportion of patients who have managed to survive beyond that point in time. The @tbl-tb2 presents at each time $t_j$ in which an event occurred, the total number of patients at risk ($r_j$) just before the $t_j$, the number of events at that time ($d_j$), the conditional probability and the cumulative probability of surviving with the standard error and 95% confidence interval (lower 95% CI, upper 95% CI).

```{r}
#| message: false
#| warning: false
#| label: tbl-tb2
#| tbl-cap: "sdf sfsf"


tb.km.A <- data.frame(tb.A, SE = round(km.A$std.err, 3), LCL = round(km.A$lower, 3), UCL = round(km.A$upper, 3))

tb.km.A |> 
  select(-n.censor) |> 
  filter(prob.A != 1) |> 
  mutate(j = row_number()) |> 
  relocate(j) |> 
  gt() |> 
  cols_label(
    time = html("Survival times (months), t<sub>j</sub>"),
    n.risk = html("No. at risk, r<sub>j</sup>"),
    n.event = html("No. of deaths, d<sub>j</sup>"),
    prob.A = html("Conditional probability of surviving, P(t<sub>j</sub>)"),
    surv.A = html("Cumulative probability of surviving, S(t<sub>j</sub>)"),
    SE = html("Standard Error of S(t<sub>j</sub>)"),
    LCL = html("lower 95% CI"),
    UCL = html("upper 95% CI")) |>
  tab_options(column_labels.font.weight = "bold") |> 
  cols_align(align = "center") 
```

From @tbl-tb2 we can see that the standard error of the S(t) increases over time as a result of estimating the survival probability at later times with less individuals.

Kaplan--Meier analysis is usually presented as a survival curve in addition to tabular form. The K-M curve of the estimated cumulative probability of surviving with 95% confidence interval (lower 95% CI, upper 95% CI) is depicted in @fig-p1.

```{r}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
#| label: fig-p1
#| fig-cap: sdfsfsf

km.A |>
  ggsurvfit(linewidth = 1, color = "#077E97" ) +
  theme_prism(palette = "winter_bright", base_size = 12) + 
  add_confidence_interval(fill = "#077E97") +
  add_censor_mark(color = "brown", size = 3.5) +
  add_risktable(risktable_stats = c("n.risk", "cum.censor", "cum.event")) +
  add_quantile(y_value = 0.5, color = "gray50", linewidth = 0.75) +
  scale_x_continuous(expand = c(0.018, 0, 0.02, 0),
                     limits = c(0, 35), breaks = seq(0, 35, 5)) +
  scale_y_continuous(expand = c(0.018, 0, 0.05, 0)) +
  scale_colour_prism() +
  labs(title = "Kaplan-Meier curve for the new therapy A",
       x = "Time in months", 
       y = "Cumulative Survival Probability") +
  theme(panel.grid.major.y = element_line(linewidth = 0.02, color = "grey80"))
```

In @fig-p1, the time post randomization in months is represented on the x-axis and the cumulative survival probability is plotted on the y-axis. The K-M curve is a stepped line, rather than a smooth curve, since it is horizontal when there is no event and drops only at times when a death occurs. Additionally, the shaded region represents the 95% confidence interval of the $S(t_j)$. We observe that the uncertainty of the KM estimate is increased over time which is indicated by the **wider** confidence intervals at later times.

::: callout-tip
## Basic characteristics of the Kaplan-Meier graph

-   At time $t_{0} = 0$, all patients are at risk and hence, the cumulative probability of surviving is $S(t_{0}) = S(0) = 1$.
-   The survival curve is defined only up to 35 months, the largest of the observation times.
-   Censoring times are marked on the K-M curve as "＋" symbols.
:::

At the bottom of the graph in @fig-p1 there is a table that presents, the number of participants at risk, the **cumulative** number of censored observations, and the **cumulative** number of patients having experienced an event at specific time points (0, 5, 10, 15, 20, 25, 30 and 35 months).

We can also estimate the median survival time -the time point at which half of the patients have survived- graphically. In @fig-p1, from the mid-point of the survival axis (Y = 0.50) move horizontally until the curve is crossed, then drop vertically to the time axis (X-axis) to find the corresponding time. In our example, the median survival is approximately 23 months for the new therapy A.

::: callout-important
If the survival curve does not drop to 0.50 or below then the median survival time cannot be computed.
:::

If we want to find in R the **median time** of surviving for patients in the new therapy A, we have to call the `km.A` survival object:

```{r}
km.A
```

We observe that the median time equals to **23 months** for the patients receiving the new therapy A.

## Comparing survival curves

### Graphical comparison

Survival of two or more groups of patients can be compared graphically. In our example, we are interested in comparing the survival curves between the new therapy and the standard therapy.

-   First, in `survfit2()` function we need to replace \~1 in the model formula with the `intervention` variable (i.e Surv \~ intervention):

```{r}
# KM curves ploted by therapy
km.AB <- survfit2(Surv(time, status) ~ intervention, data = dat)
```

-   Then, we use the `ggsurvfit()` function to dispaly a Kaplan-Meier curve for each therapy (@fig-p2):

```{r}
#| message: false
#| warning: false
#| fig-height: 6.5
#| fig-width: 8
#| label: fig-p2
#| fig-cap: sdfsfsf

km.AB |>
  ggsurvfit(linewidth = 1) +
  theme_prism(palette = "winter_bright", base_size = 12) + 
  add_confidence_interval() +
  add_censor_mark(color = "brown", size = 3.5) +
  add_risktable(risktable_stats = c("n.risk", "cum.censor", "cum.event"), size = 4) +
  add_quantile(y_value = 0.5, color = "gray50", linewidth = 0.75) +
  scale_x_continuous(expand = c(0.018, 0, 0.02, 0), limits = c(0, 35), breaks = seq(0, 35, 5)) +
  scale_y_continuous(expand = c(0.018, 0, 0.05, 0)) +
  scale_color_manual(values = c("#077E97", "#800080")) +
  scale_fill_manual(values = c("#077E97", "#800080")) +
  labs(title = "Kaplan-Meier curves (new therapy A vs standard therapy B)",
       subtitle = glue::glue("Log-rank test {survfit2_p(km.AB)}"),
       x = "Time in months", 
       y = "Cumulative Survival Probability") +
  theme(panel.grid.major.y = element_line(linewidth = 0.02, color = "grey80"),
        legend.position = c(0.85, 0.89))
```

K-M plot reveals that patients receiving the new therapy have higher probability of surviving through the whole time period. Particularly, the median survival time for each group can be estimated calling the object `km.AB`:

```{r}
# call the km.AB object to find which month corresponds to median survival run
km.AB
```

The median survival of patients in the new therapy (18 months) is higher than in standard therapy (9 months).

We can plot the cumulative risk function (aka "cumulative incidence" or "cumulative events"), $F(t) = 1 - S(t)$, with argument type = "risk".

```{r}
#| message: false
#| warning: false
#| fig-height: 6.5
#| fig-width: 8
#| label: fig-p3
#| fig-cap: sdfsfsf


km.AB |>
  ggsurvfit(type = "risk", linewidth = 1) +
  theme_prism(palette = "winter_bright", base_size = 12) + 
  add_confidence_interval() +
  add_censor_mark(color = "brown", size = 3.5) +
  add_risktable(risktable_stats = c("n.risk", "cum.censor", "cum.event"), size = 4) +
  add_quantile(y_value = 0.5, color = "gray50", linewidth = 0.75) +
  scale_x_continuous(expand = c(0.018, 0, 0.02, 0), limits = c(0, 35), 
                     breaks = seq(0, 35, 5)) +
  scale_y_continuous(expand = c(0, 0.018, 0.02, 0)) +
  scale_color_manual(values = c("#077E97", "#800080")) +
  scale_fill_manual(values = c("#077E97", "#800080")) +
  labs(title = "Kaplan-Meier curves (new therapy A vs standard therapy B)",
       subtitle = glue::glue("Log-rank test {survfit2_p(km.AB)}"),
       x = "Time in months", 
       y = "Cumulative Incidence") +
  theme(panel.grid.major.y = element_line(linewidth = 0.02, color = "grey80"),
        legend.position = c(0.85, 0.89))
```

### The log-Rank test

The log-rank test is a non-parametric method for comparing two or more survival curves. It examines the distribution of the entire survival time rather than comparing the survival probability at specific time points.

The log-rank test can be used to test the global null hypothesis that the curves are the same.

::: {.callout-note icon="false"}
## Null hypothesis and alternative hypothesis

-   $H_0$: the survival curves of the two groups of patients are not different
-   $H_1$: the survival curves are different
:::

```{r}
#Run log-Rank test to compare the curves for the intervention groups A Vs B
survdiff(Surv(time, status) ~ intervention, data = dat, rho = 0)
```
